---
title: "R Learning Guide"
author: "Mike Bader"
date: "February 2018"
output:
  pdf_document:
    keep_tex: yes
    toc: true
    toc_depth: 3
  md_document:
    variant: markdown
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())

dataDIR <- "/Users/bader/work/data/dcas/dcas2016/Dataset/"
dcas_fname <- paste0(dataDIR,"DCAS_2016_weighted.csv")
dcas <- read.csv(dcas_fname)

setwd("~/work/Teaching/Courses/ResearchPracticum/RLearningGuide/")
```

Welcome to R!
=============

You are embarking on an exciting adventure to learn extremely powerful software that you can download **for free**. 

## What is R? ##

Let's start by describing this powerful tool to start. R is both:

-   A statistical computing software

-   A full-fledged programming language

If you have learned other statistical software, you might wonder why you would want to learn another. Especially because many find that R has a steeper learning curve than these software programs. 

## Difference from Other Statistical Software ##

Other statistical software like Stata, SPSS, and SAS do one thing: they
take a table (spreadsheet) of data and statistical commands to analyze
that data.

R can do this, too. But R can do many more things because R is an entire
programming language. It has many different types of data that you can
use. You can put in a table (called a “data frame” in R-speak), but you
have lots of other types of data that you can use. It offers much more
flexibility than other statistical software.

BUT, that flexibility comes at the cost of simplicity.

## Why R? ##

So why make you learn something that’s harder than what you have already
learned?

-   R is open source (free as in speech), which means that many different people can help
    extend the language to do common tasks or specialized analysis

-   R is the future of statistical software; employers will be
    increasingly moving to R

-   Lest we forget, let us remember that R is **free** (as in beer); 
    you can use it on any computer and with
    any employer. Many of you will be working for employers that might
    not have large budgets; learning free software means that you will
    always be able to use it.

      ----------- --------------- ----------------------------
      SAS[^1]          \$9,720.00 + Windows only
      SPSS[^2]      \$99.00/month + \$237.00/month add-ons
      Stata[^3]        \$1,695.00 + \$300 for large datasets
      R                    \$0.00 
      ----------- --------------- ----------------------------

[^1]: For SAS Analytics Pro
    <https://www.sas.com/en_us/software/analytics-pro.html>; January 20,
    2018

[^2]: <https://www.ibm.com/products/spss-statistics/pricing>; January
    20, 2018

[^3]: Stata 15 government & nonprofit pricing
    <https://www.stata.com/order/new/gov/single-user-licenses/dl/>;
    January 20, 2018
    
We’re talking about a substantial amount of money!

## Getting R ##

### Download R ###

I hope that I have convinced you to at least try R at this point. 
That means you will need to get this amazing, powerful, and 
cost-effective software. You will need to download R. To do so
Go to <https://www.r-project.org/> and look
for the link that says “Download.” After you do that, you will be sent
to a list of mirrors (servers that distribute R so that no single server
gets overloaded). Find the United States and download one from that
lists (I use the [National Institute for Computational Sciences in Oak
Ridge, TN](https://mirrors.nics.utk.edu/cran/)).

### RStudio ###

If you open up R after downloading it, you will see that it gives you
*nothing* except a command line to type in commands to be run. This is
not ideal and would make R *really* difficult if that was the only way
with which you could interact with R.

That is where RStudio comes in. RStudio is an *Integrated Development
Environment (IDE)* for R. That means that RStudio gives you a window
into R–basically it acts as a gopher between you and R.

In acting as a gopher, it provides a nice *graphical user interface
(GUI)* that makes R a lot easier to use.

There are four different windows in RStudio that you will use:

Source (ctrl-1)

:   This is where you keep your scripts where you do your analysis.
    These are the same as <span>`.do`</span> files in Stata.

Console (ctrl-2)

:   This is where you can type commands directly to tell R what to do.
    You should use this sparingly because only entering commands in the
    console makes it difficult to reproduce what you have done. My
    advice: work in the source window, then copy/paste into the console
    (or source the command directly by pressing ‘ctrl-Enter’—we’ll get
    to that)

Environment

:   This window shows you all of the objects that are currently
    available in R. Remember how I told you that R can hold different
    types of data? This is where you will find all of them. A second tab
    in that window contains your history, which contains the list of
    commands that you have entered into the console.

Viewer

:   This window is where you will view help files and plots that you
    make (you can view a whole bunch of other stuff down there, but
    those are for more advanced applications).

To get RStudio, you will go to <https://www.rstudio.com/> and then find
the link for RStudio and download. This software is also free.

### Getting Help ###

As you go on this adventure, you will likely need help along the way. There are a couple of great resources on the web to help you learn just about everything that you would like to do. 

[Stack Overflow](https://stackoverflow.com/questions/tagged/r) 

:   provides user-generated answers to questions posed in forums. Good answers tend to filter to the top of the list based on an ingenuous ranking system of votes and points. 

[Rseek](https://rseek.org/) 

:   gives Google-like search functionality to find R-related material. 

[R-bloggers](https://www.r-bloggers.com/) 

:   is a blog with posts that cover tutorials and code for basic to advanced material


```{r child="ch1_variables_and_data/RLearningGuide_variables_and_data.Rmd"}
```

```{r child="ch2_descriptive_statistics/ch2_descriptive_statistics.Rmd"}
```
